When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-04-06 doc/paper/vpd_et_paper.tex
done with pierre's comments up to section 3.4

\begin{abstract}
  Increasing vapor pressure deficit increases atmospheric demand for
  water, and vapor pressure deficit is expected to rise with
  increasing CO$_2$. While increased evapotranspiration (ET) in
  response to increased atmospheric demand seems intuitive, plants are
  capable of reducing ET in response to increased VPD by closing their
  stomata, in an effort to conserve water. Here we examine which
  effect dominates response to increasing VPD: atmospheric demand and
  increases in ET, or plant physiological response (stomata closure)
  and decreases in ET. We use Penman-Monteith, combined with
  semi-empirical optimal stomatal regulation theory and underlying
  water use efficiency, to develop a theoretical framework for
  understanding how ET responds to increases in VPD.

@@

The ET response to changes in VPD alters water partitioning between
the soil and atmosphere. If ecosystem plant response reduces ET with
atmospheric drying then soil moisture will be better conserved. This
would seem a sensible evolutionary strategy to cope with aridity. If
stomata were fully passive \citep [similar to soil pores,
e.g. ][]{Or_2013}, increased atmospheric aridity would strongly reduce
soil moisture \citep{Berg_2017}. In turn, this would further increase
aridity as low soil moisture levels increase the Bowen ratio, and
cause increased temperature and atmospheric drying
\citep[][]{Bouchet_1963, Morton_1965, Brutsaert_1999, Ozdogan_2006,
  Salvucci_2013, Gentine_2016, Berg_2016}. This however would not seem
to be a sensible strategy for plants from an evolutionary
standpoint. As a counterpoint, one may argue that increases in ET with
increasing VPD could increase the likelihood of precipitation
\citep[e.g.,][]{Findell_2011}. However, increases in ET do not always
guarantee an increased likelihood of precipitation, which depending on
environmental conditions could cause a decrease in the likelihood
of precipitation \citep[][]{Gentine_2013}. Furthermore, any increases
in precipitation are likely to be non-local, such that plants giving
up water to the atmosphere are not guaranteed to reap the benefits of
water returned from the atmosphere to the soil. Using this logic, from
an ecosystem evolutionary perspective, water stored in soil is worth
much more than the chance of water returned as precipitation.

@@

Recent results have drastically improved our understanding of VPD's
impact on physiology. \citet{MEDLYN_2011} developed a model for
leaf-scale stomatal conductance ($g_s$), including VPD response, by
combining an optimal photosynthesis theory \citep{Farquhar_1980,
  Katul_2010} with an empirical approach, and extended use of this
model to the ecosystem scale in \citet{Medlyn_2017}. Additionally,
\citet{Zhou_2014} demonstrated that a quantity underlying water use
efficiency $\left(uWUE = \frac{GPP\; \sqrt{VPD}}{ET}\right)$ properly
captures a constant relationship between GPP, ET, and VPD over a
diurnal cycle. uWUE is also remarkably well conserved in the growing
season across space and time, within a PFT \citep{Zhou_2015}. These
results provide novel tools to examine how ET varies in response to
VPD, while explicitly including VPD's effect on stomatal conductance
for the first time. In this manuscript, we leverage these tools to
derive the theoretical response of ET to VPD using a Penman Monteith
framework. Our theory is validated and tested at multiple
eddy-covariance stations spanning various climates and plant
functional types.

@@

where GPP is the ecosystem scale gross primary production, and $g_1$
is an ecosystem scale analogue to $g_{1-leaf}$. We solve for $g_1$
following \citet{Medlyn_2017} (Equation 5), and take the median $g_1$
value to be representative of each PFT (Table \ref{pft}).

@@

is relatively constant across time and moisture conditions within a
plant functional type, and correctly captures a constant relationship
between GPP, ET and VPD over a diurnal cycle \citep{Zhou_2014}. We can
use uWUE to remove the $GPP$ dependence from $g_s$ in a way that makes
the PM equation analytically tractable:

@@

With Equation \ref{et} we have derived for the first time, using
recent results \citep[][]{MEDLYN_2011, Zhou_2014, Zhou_2015,
  Medlyn_2017}, ET explicitly as function of environmental variables
and two plant-specific constants, the slope parameter ($g_1$), and
uWUE, both reflecting water conservation strategy. The slope parameter
is related to the willingness of stomata to trade water for CO$_2$ and
to keep stomata open. uWUE is a semi-empirical ecosystem-scale
constant related to how WUE changes with VPD (specifically
$VPD^{-1/2}$). It is also roughly proportional to physical constants:

@@

This derivation relied either implicitly or explicitly on several
assumptions. First, the underlying theory used to derive Equations
\ref{leaf_medlyn} and \ref{medlyn} rely on an implicit assumption that
$ET=g_s \cdot VPD$; physically this implies that leaves are perfectly
coupled to the atmosphere. In reality, for some conditions and plant
types the leaves can become decoupled from the boundary layer
\citep{De_2017, Medlyn_2017}. Given our focus on the growing season,
which is usually characterized by relatively high insolation inducing
instability and convective boundary layers, we would
expect the surface to be generally well coupled. Additionally, by following
\citet{Medlyn_2017}'s method for calculating ecosystem scale $g_1$
using PM,
we have stronger confidence that surface coupling is represented
within our framework, even if it is not represented in the underlying
theory used to develop stomatal conductance ($g_s$). An additional
assumption in the formulation of $uWUE$
\citep{Zhou_2014, Zhou_2015} and
\citet{Medlyn_2017}'s stomatal conductance model is that
direct soil evaporation (E) contributions to ET remain
small relative to transpiration (T). This should be more true during the growing season. However, some plant types allow for systematically larger
contributions of ``E'' in ET, particularly those with less dense
canopies and smaller absolute amounts of transpiration. We therefore
might expect that the theory will be most applicable to forest PFTs,
which will be most strongly coupled to the boundary layer due to
larger surface roughness, and will also generally have the highest
ratios of transpiration to evaporation. Finally, for the goal of
developing PFT-wide intuition, we assume that $g_1$ and $uWUE$ are
constant in space and time. Both quantities have been shown to be well
conserved over space and time within a PFT relative to inter-PFT
variability. However, we would expect them to vary somewhat with
environmental conditions including soil water content, temperature, and
with intra-PFT plant-specific characteristics like wood density for
tree PFTs \citep{Lin_2015}. Given these strong assumptions made with
the goal of understanding broad, leading order plant behaviors, we
take the extra care of including and analyzing uncertainty (Section
\ref{testing}) and examining sensitivity to the exponent of the VPD
response, which is derived using assumptions that $ET=g_s \cdot VPD$ (Section
\ref{functional_form}).

We note one final comment on our derivation which is relevant for
drought indices. If we approximate $c_a$ at a global mean CO$_2$
concentration, then the RHS of Equation \ref{et} is fully defined
using commonly available weather station data and the constants
published here. This makes Equation \ref{et} a useful alternative to
PET in drought indices and hydrometeorological analysis for vegetated
surfaces. Equation \ref{et} better reflects the physics of water
exchange at the land surface and would only require fitting of uWUE
and $g_1$, and can also account for changes in CO$_2$ concentration
(assuming some relationship between $\lambda$ and [CO$_2$]), contrary
to typical drought indices such as the Palmer Drought Severity Index
(PDSI) \citep{Swann_2016}. Lastly, all code and data used in this
analysis, including those used to generate the figures and tables, are
publicly available at https://github.com/massma/climate\_et.

@@

Regarding the frequency of negative and positive ET response, all PFTs
exhibit a decreasing ET response with increasing VPD (physiologically
controlled, water conservative response) for the majority of
observations. The more water conservative PFTs generally exhibit
higher frequency of negative ET response, especially when one factors
in the distribution of environmental VPD (e.g. SAV and WSA grow in
more arid climates, Figure \ref{idealized_sign}). In general the bulk
statistics match our theoretical expectations well, with the caveat
that inclusion of uncertainty shifts crops towards a slightly more
negative response, and shifts many of the other PFTs, which still
exhibit a high frequency of negative ET response, towards more frequent
occurrence of positive ET response than the theory and Figure
\ref{idealized_sign} would suggest. The bulk statistics motivate a
more thorough examination of the structure of uncertainty and more
sophisticated validation of our theory's performance against
observations.

>>>>

\begin{abstract}
  Increasing vapor pressure deficit increases atmospheric demand for
  water, and vapor pressure deficit is expected to rise with
  increasing greenhouse gases. While increased evapotranspiration (ET) in
  response to increased atmospheric demand seems intuitive, plants are
  capable of reducing ET in response to increased VPD by closing their
  stomata, in an effort to conserve water. Here we examine which
  effect dominates response to increasing VPD: atmospheric demand and
  increases in ET, or plant physiological response (stomata closure)
  and decreases in ET. We use Penman-Monteith, combined with
  semi-empirical optimal stomatal regulation theory and underlying
  water use efficiency, to develop a theoretical framework for
  understanding how ET responds to increases in VPD.

@@

The ET response to changes in VPD alters water partitioning between
the soil and atmosphere. If ecosystem plant response reduces ET with
atmospheric drying then soil moisture will be better conserved. This
would seem a sensible evolutionary strategy to cope with aridity. If
stomata were fully passive \citep [similar to soil pores,
e.g. ][]{Or_2013}, increased atmospheric aridity would strongly reduce
soil moisture \citep{Berg_2017}. In turn, this would further increase
aridity as low soil moisture levels increase the Bowen ratio, and
cause increased temperature and atmospheric drying
\citep[][]{Bouchet_1963, Morton_1965, Brutsaert_1999, Ozdogan_2006,
  Salvucci_2013, Gentine_2016, Berg_2016}. This however would not seem
to be a sensible strategy for plants from an evolutionary
standpoint. As a counterpoint, one may argue that increases in ET with
increasing VPD could increase the likelihood of precipitation
\citep[e.g.,][]{Findell_2011}. However, increases in ET do not always
guarantee an increased likelihood of precipitation, which depending on
environmental conditions could cause a decrease in the likelihood of
precipitation \citep[][]{Gentine_2013}. Furthermore, any increases in
precipitation are likely to be non-local, such that plants giving up
water to the atmosphere are not guaranteed to reap the benefits of
water returned from the atmosphere to the soil. Using this subjective
logic, from an ecosystem evolutionary perspective, water stored in
soil seems to be worth much more than the chance of water returned as
precipitation.

@@

Recent results have drastically improved our understanding of VPD's
impact on physiology, especially at the leaf
level. \citet{MEDLYN_2011} developed a model for leaf-scale stomatal
conductance ($g_s$), including VPD response, by combining an optimal
photosynthesis theory \citep{Farquhar_1980, Katul_2010} with an
empirical approach, and extended use of this model to the ecosystem
scale in \citet{Medlyn_2017}. Additionally, \citet{Zhou_2014}
demonstrated that a quantity underlying water use efficiency
$\left(uWUE = \frac{GPP\; \sqrt{VPD}}{ET}\right)$ properly captures a
constant relationship between GPP, ET, and VPD over a diurnal cycle at
the ecosystem scale. uWUE is also remarkably well conserved in the
growing season across space and time, within a PFT
\citep{Zhou_2015}. These results provide novel tools to examine how ET
varies in response to VPD, while explicitly including VPD's effect on
stomatal conductance for the first time. In this manuscript, we
leverage these tools to derive the theoretical one-way response of ET
to VPD with other environmental variables properly controlled for,
i.e. we take the partial derivative of ET with respect to VPD. Our
theory is validated and tested at multiple eddy-covariance stations
spanning various climates and plant functional types.

@@

where GPP is the ecosystem scale gross primary production, and $g_1$
is an ecosystem scale analogue to $g_{1-leaf}$. We solve for $g_1$
following \citet{Medlyn_2017} (Equation 5), and take the median $g_1$
value to be representative of each PFT instead of the mean to avoid extra weighting of rare outliers (Table \ref{pft}).

@@

is relatively constant across time and moisture conditions within a
plant functional type, and correctly captures a constant relationship
between GPP, ET and VPD over a diurnal cycle \citep{Zhou_2014}. The
theoretical derivation of the square root VPD dependence in $uWUE$
leverages the same asumptions used in \cite{MEDLYN_2011} to derive the
square-root VPD dependence of the stomatal conductace model (Equation
\ref{medlyn}).  We can use uWUE to remove the $GPP$ dependence from
$g_s$ in a way that makes the PM equation analytically tractable:

@@

By accounting for photosythesis changes in ecosystem conductace, with
Equation \ref{et} we have derived for the first time, using recent
results \citep[][]{MEDLYN_2011, Zhou_2014, Zhou_2015, Medlyn_2017}, ET
explicitly as function of environmental variables and two
plant-specific constants, the slope parameter ($g_1$), and uWUE, both
reflecting water conservation strategy. The slope parameter is related
to the willingness of stomata to trade water for CO$_2$ and to keep
stomata open. uWUE is a semi-empirical ecosystem-scale constant
related to how WUE changes with VPD (specifically $VPD^{-1/2}$). It is
also roughly proportional to physical constants:

@@

This derivation relied either implicitly or explicitly on several
assumptions. First, we assume that VPD at the leaf surface is the same as VPD at measurement height; physically this implies that leaves are perfectly
coupled to the atmosphere. In reality, for some conditions and plant
types the leaves can become decoupled from the boundary layer
\citep{De_2017, Medlyn_2017}. Given our focus on the growing season,
which is usually characterized by relatively high insolation inducing
instability and convective boundary layers, we would expect the
surface to be generally well coupled. An
additional assumption in the formulation of $uWUE$ \citep{Zhou_2014,
  Zhou_2015} and \citet{Medlyn_2017}'s stomatal conductance model is
that direct soil evaporation (E) contributions to ET remain small
relative to transpiration (T). This should be more true during the
growing season. The ratio of E to T may increase immediately after
rainfall events due to high soil moisture and ponding, but VPD is
generally low anyways during these times. However, some plant types
allow for systematically larger contributions of E in ET,
particularly those with sparse canopies and smaller relative amounts
of transpiration. We therefore might expect that the theory will be
most applicable to forest PFTs, which will be most strongly coupled to
the boundary layer due to larger surface roughness, and will also
generally have the highest ratios of transpiration to
evaporation. Finally, for the goal of developing PFT-wide intuition,
we assume that $g_1$ and $uWUE$ are constant in space and time. Both
quantities have been shown to be well conserved over space and time
within a PFT relative to inter-PFT variability. However, we would
expect them to vary somewhat with environmental conditions including very low
soil water content, temperature, and with intra-PFT plant-specific
characteristics like wood density for tree PFTs
\citep{Lin_2015}. Given these strong assumptions made with the goal of
understanding broad, leading order plant behaviors, we take the extra
care of including and analyzing uncertainty (Section \ref{testing})
and examining sensitivity to the exponent of the VPD response (Section
\ref{functional_form}).

We note one final comment on our derivation which is relevant for
drought indices. If we approximate $c_a$ at a global mean CO$_2$
concentration, then the RHS of Equation \ref{et} is fully defined
using commonly available weather station data and the constants
published here. This makes Equation \ref{et} a useful alternative to
PET in drought indices and hydrometeorological analysis for vegetated
surfaces. Equation \ref{et} better reflects the physics of water
exchange at the land surface and would only require fitting of uWUE
and $g_1$, and can also account for changes in CO$_2$ concentration
(assuming some relationship between $\lambda$ and [CO$_2$]), contrary
to typical drought indices such as the Palmer Drought Severity Index
(PDSI) \citep{Swann_2016, Lemordant_2016, Lemordant_2018}. Lastly, all
code and data used in this analysis, including those used to generate
the figures and tables, are publicly available at
https://github.com/massma/climate\_et.

@@

Regarding the frequency of negative and positive ET response, all PFTs
exhibit a decreasing ET response with increasing VPD (physiologically
controlled, water conservative response) for the majority of
observations. The more water conservative PFTs generally exhibit
higher frequency of negative ET response, especially when one factors
in the distribution of environmental VPD (e.g. SAV and WSA grow in
more arid climates, Figure \ref{idealized_sign}). In general the bulk
statistics match our theoretical expectations well, with the caveat
that inclusion of uncertainty shifts crops towards a slightly more
negative response to VPD, and shifts many of the other PFTs, which still
exhibit a high frequency of negative ET response to VPD, towards more frequent
occurrence of positive ET response than the theory and Figure
\ref{idealized_sign} would suggest. The bulk statistics motivate a
more thorough examination of the structure of uncertainty and more
sophisticated validation of our theory's performance against
observations.
